A comparison of design-based and model-based approaches for spatial data.

### 2021-06-09 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
minor figure update and chunk option fixes.

```{r}
library(tidyverse)
library(here)

@@

library(cowplot)
d1 <- ggplot(data = pop_df, aes(x = x, y = y)) +
  geom_point(aes(colour = response), size = 4) +
  scale_colour_viridis_c() +
  geom_point(data = samp_df, size = 5.5, shape = 1, 
             show.legend = FALSE, stroke = 1.5) +
  theme_minimal_grid() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Design 1")
d1

>>>>

```{r, echo = FALSE, message = FALSE, warning = FALSE, out.width = "100%"}
library(tidyverse)
library(here)

@@

library(cowplot)
d1 <- ggplot(data = pop_df, aes(x = x, y = y)) +
  geom_point(aes(colour = response), size = 4) +
  scale_colour_viridis_c() +
  geom_point(data = samp_df, size = 5.5, shape = 1, 
             show.legend = FALSE, stroke = 1.5) +
  theme_minimal_grid() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Design 1")

### 2021-06-09 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
minor figure update and chunk option fixes.

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A comparison of model-based and design-based approaches for spatial data.},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{A comparison of model-based and design-based approaches for
spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matthew Higham\corref{1}}
   \ead{mhigham@stlaw.edu} 
    \author[OSU]{Lisa Madsen}

@@

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as variables measured at specific
geographic locations. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered, but
they tend to compare design-based approaches that ignore spatial
location in sampling to model-based approaches (Brus and De Gruijter,
1997; Ver Hoef, 2002; Ver Hoef, 2008). More recent overviews include
Brus (2020) and (\textbf{wang2012spatial?}), but no numerical comparison
has been made between design-based approaches that incorporate spatial
location in sampling and model-based approaches.

The rest of this paper is organized as follows. In Section
\ref{sec:background}, we compare sampling and estimation procedures
between the design-based approach and the model-based approach. In
Section \ref{sec:numstudy}, we use simulated and real data to compare
the properties of parameter estimates from both aproaches. And in
Section \ref{sec:discussion}, we end with a discussion and provide
directions for future research.

@@

\hypertarget{design-based-philosphy}{%
\subsection{Design-Based Philosphy}\label{design-based-philosphy}}

Design-Based Overview

The design-based approach to inference uses characteristics of the
sampling design to estimate parameters of interest. The design-based
approach assumes the data, \(\mathbf{Z}\), are fixed. Randomness is
incorporated through the random selection of sampling units based on
probabilities derived using the sampling design itself. Some examples of
commonly used sampling designs include independent random sampling
(IRS), stratified random sampling, and cluster sampling. Viewing the
data as fixed and incorporating randomness through the sampling design
yields estimators having very few assumptions. Asymptotic properties of
these types of estimators are derived using limiting arguments. Means
and totals, for example, are asymptotically Gaussian. Särndal et al.
(2003) and Lohr (2009) provide thorough reviews of the design-based
approach.

\hypertarget{model-based-philosphy}{%
\subsection{Model-Based Philosphy}\label{model-based-philosphy}}

Model-Based Overview

On the other hand, model-based inference imposes additional assumptions
on the data with a potential to provide more precise estimators if the
additional assumptions hold. Instead of estimating true but unknown
parameters, the goal of model-based inference in the spatial context is
often \emph{prediction} of an unknown quantity. This is a fundamental
philosophical difference between sampling-based and model-based
approaches. Instead of \emph{estimating} a fixed unknown mean, we are
\emph{predicting} the value of the mean, a random variable. We know that
if we sampled all sites, we would have an exact prediction for the mean
of our one realized spatial surface, without any uncertainty. But, the
true mean of the spatial process that generated our realized data is
still not known, and, in the prediction context, we typically do not
care much about what value the mean of the underlying process takes.

Figure 1a. Data is fixed. In a finite population example, show a 3d
surface that can be generated by anything. If we repeatedly sample the
surface, then 95\% of all 95\% CIs will contain the true mean, which
never changes.

@@

Several comparisons between the two paradigms have been made in many
different contexts. Sterba (2009) give some history of the two paradigms
as well as some applications in psychological contexts. Cooper (2006)
review the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g. Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

@@

\begin{itemize}
\item
  both good: correctly specified model with high correlation
\item
  break model: non-gaussian errors
\item
  break design: small area estimation
\item
  both good?: misspecified covariance model with high correlation
\item
  break both? non-gaussian areas with smaller sample size
\item
  change n or sampling fraction
\item
  model-based: how should sample be drawn?
\end{itemize}

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A comparison of design-based and model-based approaches for spatial data.},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{A comparison of design-based and model-based approaches for
spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matthew Higham\corref{1}}
   \ead{mhigham@stlaw.edu} 
    \author[OSU]{Lisa Madsen}

@@

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as variables measured at specific
geographic locations. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered, but
they tend to compare design-based approaches that ignore spatial
location in sampling to model-based approaches (Brus and De Gruijter,
1997; Ver Hoef, 2002; Ver Hoef, 2008). More recent overviews include
Brus (2020) and Wang et al. (2012), but no numerical comparison has been
made between design-based approaches that incorporate spatial location
in sampling and model-based approaches.

The rest of this paper is organized as follows. In Section
\ref{sec:background}, we compare sampling and estimation procedures
between the design-based approach and the model-based approach. In
Section \ref{sec:numstudy}, we use simulated and real data to study the
properties of parameter estimates from both approaches. And in Section
\ref{sec:discussion}, we end with a discussion and provide directions
for future research.

@@

The design-based and model-based approaches incorporate randomness in
fundamentally different ways. In this section, we describe the role of
randomness and its effects on subsequent inferences. We then discuss
specific inference methods for the design-based and model-based
approaches for spatial data.

\hypertarget{design-based-philosphy}{%
\subsubsection{Design-Based Philosphy}\label{design-based-philosphy}}

The design-based approach assumes the data are fixed. Randomness is
incorporated in the selection of population units according to a
sampling design. A sampling design assigns a positive probability of
inclusion in the sample (inclusion probability) to each population unit.
Some examples of commonly used sampling designs include independent
random sampling (IRS), stratified random sampling, and cluster sampling.
The goal is to use the sampling design and the sampled data to estimate
population parameters like means and totals. These population parameters
are typically assumed to be fixed but unknown.

Treating the data as fixed and incorporating randomness through the
sampling design yields estimators having very few other assumptions.
Confidence intervals for these types of estimators are typically derived
using limiting arguments. Means and totals, for example, are
asymptotically normally distributed by the Central Limit Theorem.
Särndal et al. (2003) and Lohr (2009) provide thorough reviews of the
design-based approach.

\hypertarget{model-based-philosphy}{%
\subsubsection{Model-Based Philosphy}\label{model-based-philosphy}}

The model-based approach assumes the data are a random realization of a
process. Randomness is often incorporated through distributional
assumptions on the data-generating process. Instead of estimating fixed
but unknown parameters (as in the design-based approach), the goal of
model-based inference in the spatial context is often \emph{prediction}
of an unknown quantity. For example, suppose the realized mean of all
population units is the quantity of interest. Instead of
\emph{estimating} a fixed unknown mean, we are \emph{predicting} the
value of the mean, a random variable. We know that if we sampled all
population units, we would have an exact prediction for the mean of our
one realized process, without any uncertainty. But the true mean of the
spatial process that generated our realized data is still not known.
When predicting the realized mean, we typically are not interested in
the underlying process' true mean.

Assuming the data is a realization of a specific data-generating process
yields predictors that are linked to distributional assumptions. These
distributional assumptions are used to derive prediction intervals. The
distributional assumptions allow the prediction intervals can be more
precise. (\textbf{Cressie2015statistics?}) and
(\textbf{Schabenberger2017statistical?}) provide reviews of model-based
approaches for spatial data.

Several comparisons between the model-based paradigm and the
design-based paradigm have been made in many different contexts. Sterba
(2009) give some history of the two paradigms as well as some
applications in psychological contexts. Cooper (2006) review the two
approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g. Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

Figure 1a. Brus (2020): Data is fixed. In a finite population example,
show a 3d surface that can be generated by anything. If we repeatedly
sample the surface, then 95\% of all 95\% CIs will contain the true
mean, which never changes.

\includegraphics[width=1\linewidth]{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-1-1}

@@

@@

\begin{itemize}
\item
  both good: correctly specified model with high correlation
\item
  break model: non-gaussian errors
\item
  break design: small area estimation
\item
  both good?: misspecified covariance model with high correlation
\item
  break both? non-gaussian areas with smaller sample size
\item
  change n or sampling fraction
\item
  model-based: how should sample be drawn? should locations be fixed?
\end{itemize}

@@

\leavevmode\hypertarget{ref-wang2012review}{}%
Wang, J.-F., Stein, A., Gao, B.-B., Ge, Y., 2012. A review of spatial
sampling. Spatial Statistics 2, 1--14.
